Molecular transport in human pial perivasculature

### 2024-11-12 paper/supplementary_main.tex
Stopped for today.
Former-commit-id: e8071970fb905a277e20eef1c778c57f560e9cb6

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(see also \Cref{fig:intracranial_domains}). The interface between the
parenchyma and CSF spaces is given by $\partial \Omega_{\rm PAR} \cap
\partial \Omega_{\rm CSF}$ and we separate it as two parts: the
surface of the lateral ventricles $\Gamma_{\rm LV}$ and the remaining
pial interface $\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF}
\bigcap \partial \Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm
  pia}$). The remaining, outer, boundary of $\Omega_{\rm CSF}$ is
again separated into two parts: $\Gamma_{\rm SSAS}$ represents the
lower interface towards the spinal subarachnoid space (SSAS), while
$\Gamma_{\rm AM}$ is the outer interface towards the arachnoid and
dura membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower
and upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The
boundary of the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given
by $\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$ .

%% We denote the domain defined by the connected centerlines by
%% $\Lambda$.The vascular domain is then represented as the union of
%% cylindrical vessels of radius $R_1^i$ surrounding the centerlines
%% $\Lambda^i$. Moreover, we consider the surrounding perivascular spaces
%% as the union of annular cylinders of inner radius $R_1^i$ and outer
%% radius $R_2^i = \beta R_1^i$ of width $R_2^i - R_1^i = (\beta - 1)
%% R_1^i$.

@@

In this section, we describe the mathematical model~\eqref{eq:multi_transport} and associated conditions in further detail.

Recall that the 3D concentration $c$ is defined locally in each subdomain of $\Omega$ and satisfies
\begin{equation}
     \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma = f \quad \quad \mathrm{in} \quad \Omega_{\rm CSF} \cup \Omega_{\rm{PAR}}.  \label{eq:3d_pde}
\end{equation}
The above is supplemented with the following boundary conditions. 
\begin{alignat}{2}
(D \nabla (\phi c) - \bm u c ) \cdot \bm{n}_{\rm SAS} &= g_{\mathrm{influx}} &&  \quad  \mathrm{on} \quad \spinal  \\ 
% (-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\spinal} &= 0 &&  \quad  \mathrm{on} \quad \spinal^{\mathrm{out}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   \beta c  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{u}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   0  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{l}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} &= - (-D \nabla  c_{\rm{CSF}} + \bm u_{\rm CSF} c_{\rm CSF} ) \cdot \bm{n}_{\rm{Pia}} &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} & = \beta_{\rm{pia}} (c_{\rm PAR} - c_{\mathrm{CSF}}) &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}   .  
\end{alignat}

% \mar{I think we can simplify the BCs here a bit, mostly since $\bm u \cdot \bm n$ is zero everywhere except for $\Gamma_{\rm LV}$ and $\Gamma_{\rm skull}$. So I suggest we remove all terms related to this? And we probably also don't need to split $\Gamma_{SC}$ in $\Gamma_{SC}^{in}$ and  $\Gamma_{SC}^{out}$?}

@@

Here $\phi$ is the fluid volume fraction (or porosity) of the brain (where $\phi \ll 1$), of the SAS (where $\phi = 1$) and of the PVS (where $0 \ll \phi \leq 1$); $D$ is the effective diffusion coefficient of the relevant solute in the respective media~\cite{sykova2008diffusion} which takes different values over the SAS and brain,  $\bm u$ and $\hat u$ are convective velocity fields in $\Omega$ and $\Lambda$ respectively, and $\xi$ is a transfer coefficient across the PVS-CSF and the PVS-Brain interface

The notation $\overline{c}$ denotes the lateral average of the concentration over the outer perivascular boundary \footnote{
The lateral average is defined as 
\[ \overline{c}(s) = \frac{1}{A(s)} \int_{\Theta(s)} c ,\]
where $\Theta(s)$ is a cross-section of the PVS at the point on $\bm{\lambda}(s)$ on $\Lambda$ where $\bm \lambda$ is a given parameterization of $\Lambda$ and $A(s)$ is the area of $\Theta(s)$. 
}. The (Dirac delta) term $\delta_\Gamma$ is a source term concentrated on the outer boundary of the PVS (the union of the annular domains surrounding $\Lambda^i$) that models the solute exchange between the PVSs and their surroundings \footnote{The term $\delta_{\Gamma}$ is given for any function $v$, see \Cref{sec:details_numerical_method} for more details
\[\xi  (\overline{c} - \hat c ) \delta_\Gamma (v) 
= \int_{\Lambda} P \xi ( \overline{c} - \hat c ) \overline{v} 
\]
}.

\[
%\phi =  \begin{cases}
%   1  & \mathrm{in} \;  \Omega_{\rm{csf}} \\ 
%   \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
%   \end{cases}, \; 
%D= \begin{cases}
%   D_{\rm{csf}} & \mathrm{in} \;  \Omega_{\rm{csf}} \\ 
%   D_{\rm{PAR}} (1+ R) & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
%   \quad \bm u  = \begin{cases}
%   \bm{u}_{\rm{csf}} & \mathrm{in} \; \Omega_{\rm{csf}} \\ 
%   \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
%\end{cases}, \; 
\xi = \begin{cases}
\xi_{\rm{PVS}, \rm{PAR}} & \mathrm{if} |\Lambda^i \cap \Omega_{\rm{PAR}}| \neq 0 \\ 
 \xi_{\rm{PVS}, \rm{csf}} & \mathrm{if} |\Lambda^i \cap \Omega_{\rm{csf}}| \neq 0
\end{cases} . 
\]
In the above,  $|\Lambda^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Lambda^i \cap \Omega_{\rm{csf}}|$) is nonzero whenever the surrounding PVS around the line $\Lambda^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{csf}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\mathrm{PVS}, \rm{PAR}}$ (resp $\xi_{\mathrm{PVS}, \rm{CSF}}$) if the PVS lies mostly ($80$ percent) in $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$) and the interaction with $\Omega_{\rm{PVS}}$ (resp. $\Omega_{\rm{CSF}}$ is ignored.

@@

We consider non-zero convection in the CSF space $u_{\rm CSF} \not = 0$ in \eqref{eq:multi_transport_3d} resulting from steady state CSF production.  Recall that $\Omega_{\mathrm{CSF}}$ represents the CSF space and that $\Gamma_{\mathrm{skull^{u/l}}}$ and $\Gamma_{\mathrm{pia}}$, $\Gamma_{\mathrm{LV}}$ represent the boundaries facing the upper and lower parts of the dura, the pia and the lateral ventricles, respectively.  Assuming CSF production in the choroid plexus located at the lateral ventricle wall, we solve the  steady state Stokes equations for the velocity and the pressure $(\bm u_{\mathrm{CSF}}, p_{\mathrm{CSF}})$ in the CSF domain: 
\begin{subequations}
    \begin{alignat}{2}
 - \mu \Delta \bm u_{\mathrm{CSF}} + \nabla p_{\rm{CSF}} & =  0 \quad && \mathrm{in} \,\,  \Omega_{\rm CSF}, \label{eq:momnetum_equation}  \\ 
 \nabla \cdot  \bm u_{\mathrm{CSF}} & = 0 \quad && \mathrm{in} \,\,   \Omega_{\rm CSF}, \label{eq:divergence_equation}  \\ 
\mu \nabla \bm u_{\mathrm{CSF}} \cdot \bm{n} -  p \bm n  &  = -R_0 ( \bm u \cdot \bm n ) \bm n\,\,   && \mathrm{on}  \label{eq:efflux_condition} \,\, \Gamma_{\mathrm{skull^u}}, \\ 
\bm u_{\mathrm{CSF}} & = 0 && \mathrm{on} \,\, \Gamma_{\rm{pia}} \cup \Gamma_{\mathrm{skull^l}} \Gamma_{\rm{SSAS}}  \\
\bm u_{\mathrm{CSF}} \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \mathrm{on} \,\, \Gamma_{\rm{LV}} .  
\end{alignat}
\end{subequations}

Here, $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the CSF viscosity, and $\bm u_{\rm in}$ is the steady CSF flow across the lateral ventricle wall, see \Cref{tab:pvs:parameters}.
Condition \eqref{eq:efflux_condition} models an efflux site with positive resistance $R_0 \geq 0$. The Stokes system is solved with the finite element software $FEniCS$, stored and then subsequently read for all the relevant solute transport simulations. See Figure~\ref{fig:csf_flow_cardiac} for a streamline visualization of the obtained velocity profile and \Cref{sec:details_numerical_method} for details on the numerical algorithm employed.

>>>>

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(see also \Cref{fig:intracranial_domains}). The coordinates in these
3D domains is given by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
of the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

In addition, we consider two sets of perivascular networks represented
as graphs: a periarterial network $\Lambda_a$ represented by the
(connected) centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a
perivenous network $\Lambda_v$ corresponding to the centerlines
$\Lambda^i_{(v)}$ of the veins. Recall that we represent the vascular
domains as the union of cylindrical vessels of radius $R_1^i$
surrounding the centerlines $\Lambda^i$. Moreover, we consider the
surrounding perivascular spaces as the union of annular cylinders
$\Omega^i$ of inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$
and thus of width $R_2^i - R_1^i$. We denote the outer lateral surface
of the periarterial and perivenous spaces by $\Gamma_a$ and $\Gamma_v$, respectively. We
omit the subscript $a$ or $v$ when referring to any such network,
vessel segment or perivascular outer surface. We assume that $\Gamma$
is parametrized by the coordinate $s$, and with a minor abuse of
notation, simply write $s \in \Lambda$ to represent the point ${\bm
  \lambda}(s)$ on $\Lambda$ corresponding to $s$.

@@

In this section, we describe the mathematical
model~\eqref{eq:multi_transport}, the associated definitions,
interface and boundary conditions in further detail. We refer to the
main text for explicit parameter values and give the general, abstract
form here. Recall that we model a concentration field $c = c(x, t)$
for $x \in \Omega$, $t > 0$ defined in $\Omega_{\rm PAR}$ and
$\Omega_{\rm CSF}$ separately. In each of these domains, $c$ satisfies
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma = f \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivacular networks $\Lambda_a, \Lambda_v$. To summarize,
\[
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\]
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$  if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$ in which case the interaction with $\Omega_{\rm{CSF}}$ (resp. $\Omega_{\rm{PAR}}$) is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

\fixme{Also in~\eqref{eq:3d_pde}, the notation $\overline{c}$ denotes a lateral average of the concentration over the outer perivascular surfaces, defined for each centerline $\Lambda^i$ and each point $s \in \Lambda^i$ by}
\begin{equation}
  \overline{c}(s) = \frac{1}{A(s)} \int_{\Theta(s)} c 
\end{equation}
\fixme{where $\Theta(s)$ is a cross-section of the PVS $\Omega^i$ at $s$ and $A(s)$ is the area of $\Theta(s)$.} For both the periarterial and perivenous networks ($\Lambda_a, \Lambda_v$ with outer PVS boundaries $\Gamma_a, \Gamma_v$), the (Dirac delta) term $\delta_\Gamma$ is a distribution concentrated on the outer lateral surfaces of the PVSs. \discuss{More precisely, it is defined in terms of its action on any sufficiently smooth function $v : \Omega \rightarrow \R$ and for each $\Lambda^i \subseteq \Lambda$ with PVS $\Omega^i$ and outer lateral surface $\Gamma^i = \partial \Omega^i_2$ by}
\begin{equation}
  %\xi (\overline{c} - \hat c ) \delta_\Gamma (v) = \int_{\Lambda} P \xi ( \overline{c} - \hat c ) \overline{v} .
  \delta_\Gamma (v) = \int_{\Gamma} v = \sum_{i} \int_{\Gamma^i} v = \sum_i \int_{\Lambda^i} P \, \overline{v} = \int_{\Lambda} P  \, \overline{v} .
\end{equation}
where $P = P(s)$ is defined as the perimeter of each cross-section $\Theta(s)$ for $s \in \Lambda^i$.

\mer{MER: Stopped here. Feel free to check the above. I thought the overline was defined as the cross-section perimeter average and not the cross-section average. This is more compatible with the definition of $\delta_{\Gamma}$ as well.}

We refer to the reference~\cite{masri2023modelling} for a complete model derivation and analysis.

The above is supplemented with the following boundary conditions. 
\begin{alignat}{2}
(D \nabla (\phi c) - \bm u c ) \cdot \bm{n}_{\rm SAS} &= g_{\mathrm{influx}} &&  \quad  \mathrm{on} \quad \spinal  \\ 
% (-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\spinal} &= 0 &&  \quad  \mathrm{on} \quad \spinal^{\mathrm{out}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   \beta c  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{u}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   0  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{l}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} &= - (-D \nabla  c_{\rm{CSF}} + \bm u_{\rm CSF} c_{\rm CSF} ) \cdot \bm{n}_{\rm{Pia}} &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} & = \beta_{\rm{pia}} (c_{\rm PAR} - c_{\mathrm{CSF}}) &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}   .  
\end{alignat}

%When convenient, we also denote $c|_{\Omega_{\rm PAR}} = c_{\rm PAR}$
%and $c|_{\Omega_{\rm CSF}} = c_{\rm CSF}$.

@@

@@

We consider non-zero convection in the CSF space $u_{\rm CSF} \not = 0$ in \eqref{eq:multi_transport_3d} resulting from steady state CSF production.  Recall that $\Omega_{\mathrm{CSF}}$ represents the CSF space and that $\Gamma_{\mathrm{skull^{u/l}}}$ and $\Gamma_{\mathrm{pia}}$, $\Gamma_{\mathrm{LV}}$ represent the boundaries facing the upper and lower parts of the dura, the pia and the lateral ventricles, respectively.  Assuming CSF production in the choroid plexus located at the lateral ventricle wall, we solve the  steady state Stokes equations for the velocity and the pressure $(\bm u_{\mathrm{CSF}}, p_{\mathrm{CSF}})$ in the CSF domain: 
\begin{subequations}
    \begin{alignat}{2}
 - \mu \Delta \bm u_{\mathrm{CSF}} + \nabla p_{\rm{CSF}} & =  0 \quad && \mathrm{in} \,\,  \Omega_{\rm CSF}, \label{eq:momnetum_equation}  \\ 
 \nabla \cdot  \bm u_{\mathrm{CSF}} & = 0 \quad && \mathrm{in} \,\,   \Omega_{\rm CSF}, \label{eq:divergence_equation}  \\ 
\mu \nabla \bm u_{\mathrm{CSF}} \cdot \bm{n} -  p \bm n  &  = -R_0 ( \bm u \cdot \bm n ) \bm n\,\,   && \mathrm{on}  \label{eq:efflux_condition} \,\, \Gamma_{\mathrm{skull^u}}, \\ 
\bm u_{\mathrm{CSF}} & = 0 && \mathrm{on} \,\, \Gamma_{\rm{pia}} \cup \Gamma_{\mathrm{skull^l}} \Gamma_{\rm{SSAS}}  \\
\bm u_{\mathrm{CSF}} \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \mathrm{on} \,\, \Gamma_{\rm{LV}} .  
\end{alignat}
\end{subequations}
Here, $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the CSF viscosity, and $\bm u_{\rm in}$ is the steady CSF flow across the lateral ventricle wall, see \Cref{tab:pvs:parameters}.
Condition \eqref{eq:efflux_condition} models an efflux site with positive resistance $R_0 \geq 0$. The Stokes system is solved with the finite element software $FEniCS$, stored and then subsequently read for all the relevant solute transport simulations. See Figure~\ref{fig:csf_flow_cardiac} for a streamline visualization of the obtained velocity profile and \Cref{sec:details_numerical_method} for details on the numerical algorithm employed.
